---
title: 'MIE237'
author: "Neil Montgomery"
date: "2016-01-19"
output: 
  ioslides_presentation: 
    css: 'styles.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\Var}[1]{\text{Var}\left( #1 \right)}
\newcommand{\E}[1]{E\left( #1 \right)}
\newcommand{\Sample}[1]{#1_1,\ldots,#1_n}
\newcommand{\od}[2]{\overline #1_{#2\cdot}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}

# Inference with two independent numerical samples (9.8 and 10.5) continued...

## Alternative "design" for inference on mean differences

* By "design" I mean the specific plan to gather a sample. 

* So far the design was motivated by the problem: two populations, two means (and variances), interest in the difference between the mean. Plan: gather two independent samples.

> * That's not the only way to learn about a mean difference. 

> * The sampling plan might instead involve collecting pairs $\left((Y_{11}, Y_{21}), (Y_{12}, Y_{22}), \ldots, (Y_{1n}, Y_{2n})\right)$ (that are likely to be positively correlated.)

> * In which case the correct analysis is to consider what is effectively a single sample $Y_{d_1}, Y_{d_2},\ldots, Y_{d_n}$ with $Y_{d_i} = Y_{1i}-Y_{2i}$

## "Paired observations" (9.9 and part of 10.5)

* The model reduces to $Y_{d_i} = \mu_d + \varepsilon_i \text{ with } \varepsilon_i \sim N\left(\mu_d, \sigma_d^2\right)$.

* The analysis is then just a good ol' one sample $t$ test/interval.

* The book possibly doesn't make this clear enough: **The analysis must follow the model/sampling plan**

## Example: 9.92

"...Calcium is a required element for plants and animals. The amount taken up
and stored in plants is closely correlated to the amount present in the soil. It
was hypothesized that a fire may change the calcium levels present in the soil and thus affect the amount available to deer.

A large tract of land in the Fishburn Forest was selected for a prescribed burn.
Soil samples were taken from 12 plots of equal area just prior to the burn and
analyzed for calcium. Postburn calcium levels were analyzed from the same plots."

## Example: 9.92 (with a small dose of reality)

```{r, message=FALSE, error=TRUE}
library(rio)
burn <- import("Ex09.92.txt")
```

Hmm.

## Example: 9.92

```{r}
burn <- import("Ex09.92.txt", fread=FALSE) # Seems OK
```

```{r, echo=FALSE, message=FALSE}
library(dplyr)
burn %>% 
  mutate(calcium_diff = Preburn - Postburn) -> burn_diff

burn_diff %>% 
  summarize(n=n(), mean = mean(calcium_diff), sd = sd(calcium_diff))
```

```{r}
t.test(burn_diff$calcium_diff)
```

## Model assumptions

Just the normality assumption (on the differences) - all OK.

```{r, echo=FALSE}
library(ggplot2)
burn_diff %>% 
  ggplot(aes(sample = calcium_diff)) + stat_qq()
```

## The Wrong Analysis

```{r}
t.test(burn$Preburn, burn$Postburn)
```

## Cause of the difference
```{r, echo=FALSE}
library(tidyr)
burn %>% 
  mutate(ranked_plots = rank(Preburn, ties="r")) %>% 
  gather(key = "burn_status", value = "calcium", -ranked_plots) %>% 
  ggplot(aes(x=ranked_plots, y = calcium, color = burn_status)) + 
  geom_point() + scale_x_discrete()
```

## Extreme example
```{r, echo=FALSE}
set.seed(1)
eye_drop <- data.frame(ranked_subj = 1:20, drop1 = sort(rnorm(20, 60, 10)))
eye_drop %>% 
  mutate(drop2 = drop1 + rnorm(20, 6, 3)) -> eye_drops
eye_drops %>% 
  gather("drop", dilation_time, -ranked_subj) %>% 
  ggplot(aes(x=ranked_subj, y = dilation_time, color=drop)) + geom_point()
```

## Extreme example - correct

```{r}
t.test(eye_drops$drop1 - eye_drops$drop2)
```

## Extreme example - wrong
```{r}
t.test(eye_drops$drop1, eye_drops$drop2)
```



